%%% -*-LaTeX-*-

\chapter{Applications for a Client-Assisted Design}
\label{chap:applications}

We have seen the impact of Zero Copy on network transmission and memory bandwidth and how that 
compares to using Copy Out to transmit data over the network. The difference
in throughput is explicitly seen for smaller 128~B records, where owing to the descriptor overhead and anomalies discussed 
in Chapter~\ref{chap:modernnics}, Copy Out gives better throughput performance especially when the transmissions get larger.
We have already established that 128~B records show a more realistic depiction of record sizes in the future
and that seems to be the growing consensus in the community~\cite{fb-memcache,fb-workload}. 

At first, the throughput results from comparing Zero Copy and Copy Out for 128~B 
records might tempt us to think that it is always better to use a Copy Out get maximum throughput with smaller records.
The complications of implementation and the cost of overheads are also rampant when using small records with Zero Copy.
The bigger problem with Copy Out might get overlooked in the face of all these issues surrounding Zero Copy.
While transmitting at around 5.9~GB/s (near line rate for the Connect-X3\textregistered NIC),
we saw that the memory pressure on the system is close to 12~GB/s, which is half of what we evaluated our server's total available memory bandwidth. 
This implies that network transmission will bottleneck available memory bandwidth for doing useful work such as processing data in a hybrid or analytical workload.

On-Line Analytical Processing (OLAP) and hybrid workloads composed of transactional and analytical queries have fuelled a renewed interest in column stores~\cite{cstore,cstorevsrowstore} in recent years. 
Column stores are uniquely positioned to take advantage of the NIC because of their layout and compression benefits.
 The impact of modern NIC while designing column stores should also be discussed for completeness. 

This chapter introduces a client-assisted design with bounded-disorder for using Zero Copy with no-update-in-place row stores as well as column stores. 
We evaluate the improvement in throughput and efficiency to provide evidence to back this approach.
We also make informed opinions on how Zero Copy might perform on column store designs 
for in-memory databases with co-operation from clients for materializing data.


\section{Bw-Trees}
Chapter~\ref{chap:modernnics} showed us that Copy Out might offer better transmission throughput for a large set 
of scattered small records. It also showed the perils of using Copy Out where it consumes a lot of memory bandwidth, 
starving the server of memory bandwidth that could be utilized for doing useful work.
Zero Copy promises memory bandwidth savings in addition to eliminating CPU overhead, 
while Copy Out can saturate the NIC throughput. We wanted to extract the best of both worlds, 
and we wanted to analyze structures that can uniquely take advantage of the 
Zero Copy paradigm. This would imply an arrangement of small records where the total transmission size 
goes up, helping the transmission throughput while being able to do Zero Copy to help reduce memory footprint. For this 
to work, we would also need a way of efficient update and synchronization so that data being fetched for transmission remain 
stable throughout the transmission.

Microsoft\textregistered Research came up with a modern B-tree structure called 
the Bw-Tree~\cite{bw-tree} which is their primary indexing structure for SQL Server Hekaton~\cite{hekaton}.
Bw-Tree is a highly performant, ordered index employing lock and latch free concurrency and effective utilization of caches in 
modern multicore processors. In the following sections, we explore how a client-assisted design involving a no-update-in-place 
structure such as Bw-Trees can take advantage of Zero Copy and gain maximum throughput 
without much system impact. We discuss how its lock freedom and delta records make it 
particularly suitable to exploit Zero Copy in the following sections. We then present
our evaluation of transmissions involving small records tacked on to bigger base records
to show the gains in throughput without consuming more memory bandwidth. 

\section{Lock-free Indexing}
\label{sec:bwtree-intro}
The Bw-Tree~\cite{bw-tree} is an atomic record store designed for extreme
concurrency. It is an ordered index that supports basic record create, update,
and delete (CRUD) operations in addition to searches and range scans.  It is
fully lock-free and nonblocking and is optimized for modern multicore
workloads. It can store to flash, but is also intended for in-memory
workloads; it serves as the ordered secondary index structure for the in-memory SQL
Server Hekaton~\cite{hekaton} engine.

In many ways, the Bw-Tree is a conventional paged B-link tree,
but it also has unique characteristics that interact with network-layer
design choices. Its lock freedom, its elimination of update-in-place,
and its lazy consolidation of updated records in virtual memory give it
tremendous flexibility in how query results are transmitted by the NIC.

Records may be embedded within leaf pages, or the leaf pages may
only contain pointers to data records. When used as a secondary index,
typically leaf pages would contain pointers, since otherwise, each record would
have to be materialized twice and the copies would need to be kept consistent.

The key challenge in a lock-free structure is providing atomic reads, updates,
inserts, and deletes without ever being able to quiesce ongoing operations (not
even on portions of the tree). Bw-Tree solves this problem by eliminating
update-in-place. All mutations are written to newly allocated memory; then
the changes are installed with a single atomic compare-and-swap instruction
that publishes the change.  Figure~\ref{fig:bw-tree} shows how this works.
In place updates are avoided by creating delta records ``off to the side'' 
that describe a logical modification to a page. Delta records are
prefixed to a chain ultimately attached to a base page.  When delta chains
grow long, they are compacted together with the base page to create a new base page.
All references to pages are translated through a mapping table that maps page
numbers to virtual addresses. This allows pages to be relocated in memory, and
it allows the contents of a page to swapped with a single atomic
compare-and-swap (CAS) operation.
\section{Delta Records}
\label{sec:deltarecords}
One of the key innovations of the Bw-Tree is its use of {\em delta records},
which make updates inexpensive.
Delta records allow the Bw-Tree to logically modify the
contents of an existing page without blocking concurrent page readers, without
atomicity issues, and without recopying the entire contents of the page for
each update.  Whenever a mutation is made to a page, a small record is
allocated, and the logical operation is recorded within this delta record. The delta
record contains a pointer to the page that it logically modifies. It
is then atomically installed by performing a CAS operation on the
mapping table that re-points the virtual address for a particular page number
to the address of the delta record.

Some threads already reading the original page contents may not see
the update, but all future operations on the Bw-Tree that access that page
will see the delta record. As readers traverse the tree, they consider
the base pages to be logically augmented by their delta records. Delta records
can be chained together up to a configurable length.  When the chain becomes
too long, a new base page is formed that combines the original base page
contents with the updates from the deltas. The new page is swapped in
the same way as other updates.

Read operations that run concurrent to update operations can observe superseded
pages and delta records, so their garbage collection must be deferred.
To solve this, each thread that
accesses the tree and each unlinked object are associated with a current {\em epoch}.
The global epoch is periodically incremented. Memory for an unlinked object can be
recycled when no thread belongs to its epoch or any earlier epoch.
The epoch mechanism gives operations consistent reads of the
tree, even while concurrent updates are ongoing. However, there is a
cost; if operations take a long time, they remain active within their epoch and
prevent reclamation of memory that has been unlinked from the data structure.


\section{NIC Implications for Bw-Tree}
Lock freedom has major implications on the in-memory layout of the
Bw-Tree. Most importantly, readers (such as the NIC DMA engine) can collect a
consistent view of the tree without interference from writers, and hold that
view consistent without stalling concurrent readers or writers to the tree.  This
natural record stability fits with the Zero Copy capabilities of modern NICs;
because the NIC's DMA engine is oblivious to any locks in the database engine,
structures requiring locking for updates would have to consider the NIC to
have a non-preemptible read lock for the entire memory region until the DMA completes.
Instead of copying records ``out'' of the data structure for transmission,
records can be accessed directly by the NIC. Eliminating the explicit copy of
the records into transmit buffers can save database server CPU and memory
bandwidth.

Page consolidation can also benefit the NIC and improve performance.  Records
in the Bw-Tree are opportunistically packed into contiguous pages in virtual
memory, but the view of a page is often augmented with (potentially many)
small delta records that are scattered throughout memory.
A database might save CPU and memory bandwidth by more
aggressively deferring or even eliminating consolidation of records into
contiguous regions of memory or pages. We show in
our evaluation that highly discontinuous data can slow
transmission throughput but that aggressive consolidation is inefficient; delta
records can dramatically reduce consolidation overheads while keeping records
sufficiently contiguous to make the NIC more efficient.
Overall, we seek to answer these key questions:
\begin{myitemize}
\item
When should records be transmitted directly from a Bw-Tree? Are there cases
where explicitly copying records into a transmit buffer is preferable to Zero Copy?
\item
How aggressive should a Bw-Tree be in consolidating records to benefit individual
clients and to minimize database server load?
\item
How does Zero Copy impact state reclamation in the Bw-Tree? Might long transmit
times hinder performance by delaying garbage collection of stale records?
\end{myitemize}



% Answering these questions?
% - When to tx zero copy versus not?
% - Show tx perf CPU use.


% K: Indexes in modern databases sustain million/op/s.

% K: To support extreme concurrency sometimes lock-free [cite BwTree, ART]

% K: Range scan performance key. But raises key question? How to inexpensively
% get the data to the NIC?

% K: Simplest approach: CO results and transmit. Tradtionally this would
% have yielded three copies. One into the results buffer, one for the kernel to
% copy the results into packet buffers, and one for the NIC to DMA the data for
% transmit.

% What about Zero Copy? Several problems:
%  - atomicity
%  - garbage collection and object lifetime
%  - packaging the objects for efficient transmit.
%    - pre-package? (paged structures)
%    - use NIC DMA engine?
% Key question? What are the gains?
%  - Reduced copies -> reduced memory bandwidth use, reduced CPU time

% K: Zero copy? Then data structure must be tied into messaging layer. Can work
% with epoch-based GC, but what about long scans? Could hold back GC and stall
% tree in the limit.




\section{Evaluation}

\subsection{Extending the Delta Format to Clients}


The experiments in Chapter~\ref{chap:modernnics} consider delivering a clean, ordered set of records to the
client. That is, the server collects and transmits the precise set of records
in the correct order, either via Copy Out or Zero Copy. Another option is to
transmit base pages along with their deltas and have clients merge the results.
This approach is attractive because it organically fits with the design of the
Bw-Tree and it suits the NIC's DMA engine well.  The NIC can transmit the
large contiguous base pages (16~KB, for example) with little CPU overhead.
It also eliminates Copy Out for small records, avoiding the ceiling of transmission throughput. %(\ref{Chnics-sec:zero-copy-tput}).
Merging records on the client side is cheap; the server can even append them to
the response in a sort order that matches the record order in the base page for
efficient $O(n)$ iteration over the records that requires no copies or sorting.


\subsection{Throughput and Efficiency of Delta Records}
\label{sec:delta-tput-efficiency}
Figure~\ref{fig:deltas-tput} shows the benefits of delta records. In our
experiment, each transmission consists of a single 16~KB base page while the
number (one to thirty one) and size (128~B and 1024~B) of the delta records attached to each transmission is varied.
The NIC benefits from the large base page, and it manages to keep the network
saturated. CPU overhead ranges from less than 0.3\% when there are a few delta
records up to about 0.6\% when there are more. This offers a 30$\times$ reduction 
in CPU overhead in comparison with our evaluation of the Copy Out approach in 
Section ~\ref{sec:overhead}. Compared to Zero Copy of scattered
small records, this approach also yields a 1.6$\times$ throughput advantage;
Figure~\ref{fig:zero-copy-tput} shows that throughput peaks around 3.4~GB/s when
records are scattered, while the delta format achieves around  5.6~GB/s. The main cost 
to this approach is the cost of consolidating delta records into new base pages 
when chains grow long; we consider this overhead in more detail in the next section.

% 5740/3532
% 1.62
% = 1.6x

\subsection{Tuning Page Consolidation}
\label{sec:consolidation}


Bw-Tree and many indexing structures are paged to amortize storage
access latency, and paging can have similar benefits for network transmission
as well. However, the user-level access of modern NICs makes interacting with
them much lower-latency than storage devices. This raises the question of
whether paging is still beneficial for in-memory structures. That is, is the
cost of preemptively consolidating updates into pages worth the cost, or is it
better to transmit fine-grained scattered records via Zero Copy or Copy Out?

%\input{fig-deltas}

% Regular CPU overhead
%    copied chunkSize chunksPerMessage   busyFrac
% 1       0       128                1 0.08386966
% 2       0       128                2 0.11668905
% 3       0       128                3 0.10951036
% 4       0       128                4 0.10252412
% 5       0       128                5 0.06622680
% 6       0       128                6 0.06367224
% 7       0       128                7 0.06985951
% 8       0       128                8 0.06227349
% 9       0       128                9 0.05706500
% 10      0       128               10 0.04953507
% 11      0       128               11 0.04680207
% 12      0       128               12 0.04342779
% 13      0       128               13 0.04162006
% 14      0       128               14 0.04048418
% 15      0       128               15 0.03841897
% 16      0       128               16 0.02195241
% 17      0       128               17 0.02131298
% 18      0       128               18 0.02084331
% 19      0       128               19 0.02049131
% 20      0       128               20 0.02022781
% 21      0       128               21 0.01989223
% 22      0       128               22 0.01952930
% 23      0       128               23 0.01917201
% 24      0       128               24 0.01921045
% 25      0       128               25 0.01872136
% 26      0       128               26 0.01838027
% 27      0       128               27 0.01814930
% 28      0       128               28 0.01809671
% 29      0       128               29 0.01792050
% 30      0       128               30 0.01771566
% 31      0       128               31 0.01749623
% 32      0       128               32 0.01781422


% Delta CPU overhead
%    copied deltaSize deltasPerMessage    busyFrac
% 1       0       128                0 0.004176226
% 2       0       128                1 0.004343818
% 3       0       128                2 0.004479262
% 4       0       128                3 0.004715360
% 5       0       128                4 0.004881293
% 6       0       128                5 0.004947248
% 7       0       128                6 0.005532234
% 8       0       128                7 0.005632129
% 9       0       128                8 0.005691837
% 10      0       128                9 0.005749774
% 11      0       128               10 0.005764381
% 12      0       128               11 0.005905733
% 13      0       128               12 0.006012879
% 14      0       128               13 0.006140493
% 15      0       128               14 0.006128670
% 16      0       128               15 0.004983695
% 17      0       128               16 0.005027644
% 18      0       128               17 0.005060442
% 19      0       128               18 0.005158658
% 20      0       128               19 0.005279367
% 21      0       128               20 0.005294916
% 22      0       128               21 0.005320413
% 23      0       128               22 0.005371369
% 24      0       128               23 0.005409892
% 25      0       128               24 0.005470040
% 26      0       128               25 0.005479338
% 27      0       128               26 0.005522727
% 28      0       128               27 0.005549651
% 29      0       128               28 0.005604431
% 30      0       128               29 0.005634841
% 31      0       128               30 0.005600399
% 32      0       128               31 0.005787729



The results of our earlier experiments help answer this question.  If Copy Out
is used to gain faster transmission of small records, then the answer is
simple. Even if every update created a complete copy of the full base page, the
extra copies would still be worthwhile so long as more pages are read per
second than updated. This is true for most update/lookup workloads and read-only
range queries make this an even better trade-off.

However, consolidation must be more conservative when using Zero Copy to yield
savings, since Zero Copy can collect scattered records with less overhead than
Copy Out. Yet there is a good reason to be optimistic.  Delta records
reduce the cost of updates for paged structures. If each base page is limited
to $d$ delta records before consolidation, the number of consolidations is
$\frac{1}{d}$. This means that allowing short delta chains dramatically reduces
consolidation costs, while longer chains offer decreasing returns. This fits with
the NIC's preference for short gather lists; allowing delta chains of length 4
would reduce consolidation by 75\% while maintaining transmit throughput that
meets or exceeds on-the-fly Copy Out.  The large 16~KB base pages also
improve transmit performance slightly, which improves efficiency.

For small records, transmitting compacted base pages that have a few deltas gives
a total CPU savings of about 10\%. For the same CPU cost, a server can perform
about 6.5~GB/s of page compaction or about 425,000 compactions per second for
16~KB pages. Even in the worst case scenario where all updates are focused on a
single page, delta chains of length four would tolerate 1.7~million updates per
second with CPU overhead lower than Copy Out. So, delta records can give the
efficiency of Zero Copy with the performance of Copy Out.


\subsection{Impact on Memory Bandwidth}
% Copy out max: 11596
% Delta max: 6226
Figure ~\ref{fig:deltas-membw} shows the total consumed memory bandwidth while 
transmitting records. It is interesting to note that regardless of the variation in 
delta record sizes, memory bandwidth stays roughly the same. Comparing Copy Out 
transmissions of similar size, we get a 46\% reduction in the consumption of 
memory bandwidth in the system. Another interesting factor is that the transmission 
involved only a few consolidated delta records, which we showed in Section~\ref{sec:consolidation} is 
beneficial; there is no added memory bandwidth pressure if we decide to transmit smaller 
or larger records. The bigger base page gets us to peak transmission, and a limited number of 
delta records is favored for aggregate CPU costs. We have arrived at the sweet spot of 
transmitting large amounts of data in a modern NIC.


\subsection{Impact on Garbage Collection}

Using Zero Copy tightly couples the higher level database engine's record
allocation and deallocation since records submitted must remain stable until
transmission completes. Fortunately, existing mechanisms in systems that avoid
update-in-place accommodate this well, like Bw-Tree's epoch mechanism described
in Section~\ref{sec:bwtree-intro}. Normally, the Bw-Tree copies-out returned records.
While an operation and its Copy Out are ongoing, the memory that contains records
that have been unlinked or superseded cannot be reclaimed. Zero Copy
effectively defers the copy until the actual time of transmission. As a result,
transmissions completions must notify the higher level database of transmission
completion to allow reclamation. This delay results in more latent garbage and
hurts the memory utilization of the system.

In practice, this effect will be small for two reasons. First, Zero Copy adds a
transmission, which completes within a few microseconds; however, it also saves a
\memcpy~ that accounts for 1~to~2~\textmu s for a 16~KB transmission. Second, the
amount of resulting data held solely for transmission should generally be
more than compensated for by eliminating the need for explicit transmit
buffers. With Copy out, the size of the transmit buffer pool is necessarily
larger than the amount of data under active transmission, and internal
fragmentation in the transmit buffers makes this worse.


\section{Zero Copy for Column Stores}

Column-oriented in-memory databases pose unique challenges in interacting with a modern NIC. 
Column stores use contiguous memory locations to store adjacent columns. They have also been known 
to use Run-Length Encoding, Dictionary Encoding, Bit-Vector Encoding, and heavier schemes 
like Lempel-Ziv Encoding for yielding better compression, which is one of the many value 
propositions of using a column-oriented layout. 

We observed in the previous  sections that Zero Copy offers promising performance if 
we had stable entries and used Zero Copy to transmit the results while responding to a range query. 
However, the fundamental problem lies in the fact that end-users expect query processing and analytics on data 
to materialize results in row-order. It has been argued before that early materialization of data
helps performance in column-oriented databases~\cite{cstore-material}. This makes the application 
of a scatter gather list difficult, and we might have already used a \memcpy ~or an alternative 
to materialize the data. 

We argue for the possibility of a different materialization strategy with more engagement from the 
clients. If the database server were to sent filtered columns and required metadata such as the 
dictionary encoding as scatter gather entries, we could leverage Zero Copy's performance. We believe 
this approach would work even for hybrid layouts~\cite{hyrise-hybridstores} that use row-oriented or 
column-oriented formats varying the width of columns according to the type of workload. Figure ~\ref{fig:rowstore-colstore}
shows how row stores and column stores could transmit data using Zero Copy using a client-assisted design.
In practice, there are numerous challenges associated with this design:
\begin{myitemize}
\item {\em Client co-operation}; this scheme of transmitting results would need smart clients that are 
capable of materializing data given the encoding scheme and the columns. We assume that the client 
has enough resources to spend in finessing the results.
\item {\em Low selectivity in queries}; we have seen packing more data yields more savings and performance 
from Zero Copy. We also know that the compression schemes in column stores can limit transfer sizes. 
As a result, queries with low selectivity are a natural fit for Zero Copy.
\item {\em Compressed metadata}; the client-assisted design would need the encoding to be transmitted
along with the results; this would need efficient compression and lean data structures for metadata.
\end{myitemize}
There needs to be further evaluation of how we can extract maximum performance and efficiency from 
a server equipped with a modern NIC.
\newline
\input{takeaway-bwtrees}

\section{Conclusion}
Zero Copy and Copy Out provides different optimizations for scattered, small records. A hybrid approach 
that leverages the high throughput that we obtain as a result of pre-assembling scattered records and 
using Zero Copy to transmit will yield good transmission throughput without consuming a lot of memory bandwidth.
\begin{myitemize}
\item With the assumption of {\em co-operative clients} and queries with low selectivity, we can continue 
to use Zero Copy and get better results while dealing with small, scattered records that are common in in-memory databases.
\item Advanced structures such as the Bw-Tree that allow no update in place and complete lock and latch freedom 
can be exploited for their suitability with the NICs. Further evaluation is required for these to work in practice.
\item Column stores could also benefit from the hybrid approach; materializing results in row-order makes Zero Copy 
more complicated in modern NICs and one possibility is to send the encoding metadata along with the results with client-assisted 
materialization.
\end{myitemize}
\newpage
\input{fig-bw-tree}
\input{fig-deltas-tput}
\input{fig-deltas-membw}
\input{fig-rowstore-colstore}
